---
title: |
  Stress Reactivity  
  Disturbances of the Neurocardiac Axis
subtitle: "Anish Sanjay Shah"
author: |
  A thesis submitted for the degree of  
  *Master of Science in Clinical Research*   
  Laney Graduate School  
  Emory University  
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: 
  bookdown::pdf_book:
    includes:
      in_header: preamble.tex
    lot: no
    lof: no
    toc_unnumber: no
    toc_appendix: yes
    toc_depth: 3
bibliography: ["../bibliography/Neurocardiology.bib", "../bibliography/Publications.bib"]
csl: "../bibliography/american-journal-of-epidemiology.csl"
link-citations: yes
documentclass: book
classoption: openany
colorlinks: yes
graphics: yes
geometry:
- top=1in
- bottom=1in
- left=1.5in
- right=1.5in
fontsize: 11pt
mainfont: Constantia
monofont: "Source Code Pro"
monofontoptions: "Scale=0.8"
header-includes:
  - \AtBeginDocument{\frontmatter}	
  - \usepackage{float}
  - \usepackage{setspace}
  - \usepackage{longtable}
  - \usepackage{booktabs}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
  - \usepackage{graphicx}
---

```{r setup, global_options, include=FALSE}
# Knitr layout options
knitr::opts_chunk$set(
  cache = TRUE,
  eval = TRUE,
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  dpi = 300,
  dev = "png",
  options("scipen" = 0, "digits" = 3)
)
```

```{r flair_color}
library(flair)
darkblue <- "#0C2340"
gold <- "#B58500"
blue <- "#012169"
yellow <- "#F2A900"
teal <- "#487F84"
orange <- "#C35413"
red <- "#DA291C"
violet <- "#6558B1"
magenta <- "#C6007E"
lightblue <- "#008DBA"
black <- "#101820"
grey <- "#B1B3B3"
```

# Preface

# Acknowledgements

Foremost, I would like to thank my mentors and mentoring team. Dr. Shah has been a tremendous supporter of my academic growth, and has challenged me to continue to explore and learn, and create an area of academic excellence for myself. Dr. Alonso and Dr. Thames have all supported my efforts through direct mentorship and academic guidance. Dr. Quyyumi and Dr. Vaccarino have created a research environment that has laid the groundwork for these projects.

Second, I would like to thank the support staff that has allowed this research to unfold, from the EPICORE group with Nancy Murrah and Lucy Shallenberger, to the MSCR program with Cheryl Sroka and the outstanding MSCR instructors.

Finally, I would like to thank my family for their patience and support.


# Abbreviations

There are several key abbreviations that will be used throughout. They have been outlined here for reference.

```{r abbreviations}
tribble(
	~Term, ~Abbreviation,
	"HRV", "Heart Rate Variability",
	"CAD", "Coronary Artery Disease",
	"MACE", "Major Adverse Cardiovascular Events",
	"Biobank", "Emory Cardiovascular Biobank",
	"MIMS", "Myocardial Infraction and Mental Stress",
	"MIPS", "Mental Stress Ischemia Mechanisms and Prognosis Study",
	"Twins", "Emory Twins Study"
) %>%
	arrange(Term) %>%
	gt() 
```

<!--chapter:end:index.Rmd-->

\mainmatter

# (PART\*) INTRODUCTION {-}

# Overview

### Research Problem

-   Mental stress can cause changes in the brain\
-   These changes can lead to depression and psych disease\
-   Those changes lead to increased cardiovascular disease\
-   Patients that are comorbid with psych and CAD do worse clinically


# Outline

1.  Research problem
1.  Purpose of research
1.  Underlying causal mechanisms
1.  Overview of how to address problem


> "Why did he die on Tuesday and not on Monday?"
>
> --- Douglas Zipes


<!--chapter:end:dissertation/01-introduction.Rmd-->

# (PART\*) BACKGROUND {-}

# Review of the Literature

# Clinical Relevance

<!--chapter:end:dissertation/02-background.Rmd-->

# (PART\*) METHODS {-}

# Specific Aims

The response to both physiological and psychological stress can be markers of overall cardiovascular adaptability. The following aims help to assess the clinical importance of stress reactivity as measured by disturbances to the neurocardiac axis. 

1. To assess the association between myocardial ischemia and coronary perfusion on cardiac autonomic activity. 
1. To determine if cardiac autonomic activity modifies the relationship between acute and chronic psychological stress and myocardial ischemia.
1. To explore the association of cardiac autonomic activity with future major adverse cardiovascular events.

# Study Design

## Population characteristics

## Measurements

## Sample size and power considerations

# Analysis

## Descriptive analysis

## Statistical inference

<!--chapter:end:dissertation/03-methods.Rmd-->

# (PART\*) RESULTS {-}

# Clinical Characteristics

Table 1

# Myocardial Ischemia

Relationship to myocardial ischemia

# Mental Stress and Myocardial Perfusion

Psychological factors and myocardial ischemia.

# Clinical Outcomes

Long-term follow-up events based on HRV.

<!--chapter:end:dissertation/04-results.Rmd-->

# (PART\*) DISCUSSION {-}

# Major Findings

# Strengths and Limitations

# Next Steps

<!--chapter:end:dissertation/05-discussion.Rmd-->

# (PART\*) CONCLUSIONS {-}

Here are my thoughts.

<!--chapter:end:dissertation/06-conclusions.Rmd-->

# (PART\*) REFERENCES {-}

<div id="refs"></div>

<!--chapter:end:dissertation/07-references.Rmd-->

# (PART\*) APPENDIX {-}

```{r, include=FALSE, echo=FALSE}
tar_load(diagrams)
tar_load(biobank_tables)
tar_load(mims_tables)
tar_load(mims_figures)
tar_load(mims_reports)
tar_load(twins_tables)
tar_load(twins_reports)
```


# TABLES AND FIGURES

\newpage

## Clinical Overview

### Overview of Stress Reactivity {#FigureDag}

```{r, fig.height=3, fig.width=5}
diagrams$dag +
	scale_color_manual(values = c(teal, gold), na.value = grey) + 
	theme(
		legend.position = "none",
		plot.caption = element_text(hjust = 0)
  ) +
	labs(
		title = "Stress Reactivity and the Neurocardiac Axis",
		caption = "Directed acyclic graph of the relationship between neurocardiac stressors and potential effect on clinical outcomes, including major adverse cardiovascular events."
	)
```

\newpage
\bfoot

### Biobank Cohort Description {#TableOneBiobank}


```{r}
biobank_tables$one %>%
	as_gt() %>%
	tab_header(title = "Emory Cardiovascular Biobank", subtitle = "Cohort Description") %>%
	tab_source_note("A description of subjects undergoing left heart catheterization with coronary angiography, including burden of coronary artery disease. CASS = Coronary Artery Surgery Score, PHQ = Patient Health Questionnaire, BMI = Body Mass Index.") 
```

\efoot
\newpage
\bland
\bfoot

### Twin Cohorts Description {#TableOneTwins}


```{r}
twins_tables$one %>%
	as_gt() %>%
	tab_header(title = "Emory Twins Study", subtitle = "Cohort Discription") %>%
	tab_source_note("Description of the veteran twin subjects within each follow-up period. They were evaluated for clinical characteristics, including quantitative myocardial perfusion imaging. THS = Twins Heart Study, SAVEIT = Stress and Vascular Evaluation in Twins, ETSF = Emory Twins Study Follow-Up.")
```

\efoot
\eland
\newpage
\bland
\bfoot

### Mental Stress Cohorts Description {#TableOneMIMS}


```{r}
mims_tables$one %>%
	as_gt() %>%
	tab_header(title = "MIMS and MIPS", subtitle = "Cohort Discription") %>%
	tab_source_note("MSIMI = Mental Stress Induced Myocardial Ischemia; PSIMI = Physical Stress Induced Myocardial Ischemia, MIMS = Myocardial Infarction and Mental Stress, MIPS = Mental Stress Ischemia Mechanisms and Prognosis Study") 
```

\efoot
\eland
\newpage
\bland
\bfoot

### HRV in Twins Cohorts {#TableTwinsHRV}

```{r}
twins_tables$hrv %>%
	as_gt() %>%
	tab_header(title = "Description of HRV", subtitle = "Emory Twins Study") %>%
	tab_source_note("Heart rate variability is described in each of the follow-up periods. HRV = heart rate variability, Dyx = kurtosis of Poincare plot, SDNN = the standard deviation of normally conducted RR intervals, RMSSD = the root mean square of successive differences in normally conducted RR intervals, PNN50 = proportion of normally conducted RR intervals that differ by more than 50 ms divided by the total number of normally conducted RR intervals")
```

\efoot
\eland

## MYOCARDIAL ISCHEMIA

\bland
\bfoot

### Relationship Between Obstructive and Non-Obstructive Coronary Artery Disease {#TableBiobankHRV}

```{r}
biobank_tables$hrv %>%
	as_gt() %>%
	tab_header(title = "HRV and Obstructive CAD", subtitle = "Emory Cardiovascular Biobank") %>%
	tab_source_note("In patients undergoing angiography, HRV metrics were described in those with both obstructive (>70%) and nonobstructive CAD, and evaluated for differences in distribution. HRV = Heart Rate Variability, CAD = Coronary Artery Disease.")
```

\efoot
\eland
\newpage
\bland
\bfoot

### HRV by Timing of Revascularization {#TableBiobankTiming}

```{r}
biobank_tables$timing %>%
	as_gt() %>%
	tab_header(title = "HRV and Timing of Myocardial Reperfusion", subtitle = "Emory Cardiovascular Biobank") %>%
	tab_source_note("HRV was measured at the start of coronary angiography, as well as intervention. Coronary arteries with obstructive disease are reperfused using balloon angioplasty and potential stenting. HRV = Heart Rate Variability, CAD = Coronary Artery Disease.")
```

\efoot
\eland
\newpage
\bland
\bfoot

### Relationship of HRV with both Mental and Physical Stress {#TableMimsStress}


```{r}
mims_reports$stress %>%
	tab_header(title = "Myocardial Perfusion Imaging with Physical and Mental Stress", subtitle = "MIMS/MIPS Cohorts") 
```

\efoot
\eland
\newpage
\bland
\bfoot

### Quantitative Myocardial Perfusion and HRV {#TableTwinsMPI}

```{r}
twins_reports$mpi %>%
	tab_header(title = "Myocardial Perfusion Imaging and Morning HRV", subtitle = "Emory Twins Study") %>%
	tab_source_note("Relationship between abnormal MPI and CFR with HRV. HRV = heart rate variability, MPI = myocardial perfusion imaging, CFR = coronary flow reserve, LF = low frequency HRV, HF = high frequency HRV, VLF = very low frequency HRV, AC = acceleration capacity")
```

\efoot
\eland
\newpage
\bland
\bfoot

### Circadian HRV and Myocardial Perfusion {#TableTwinsCircMPI}

```{r}
twins_reports$ischemia_cosinor %>%
	tab_header(title = "Circadian HRV and Myocardial Perfusion Abnormalities", subtitle = "Emory Twins Study") %>%
	tab_source_note("Myocardial perfusion was quantified as a ccontinuous variable and as a binary of abnormal or normal. The HRV metrics are measured over 24 hours using cosinor statistics. MPI = myocardial perfusion imaging, CFR = coronary flow reserve, HRV = heart rate variability, LF = low frequency HRV, HF = high frequency HRV, VLF = very low frequency HRV, AC = acceleration capacity, MESOR = midline estimating statistic of rhythm, Amplitude = maximum distance from MESOR, Phi = shift of acrophase")
```

\efoot
\eland

## MENTAL STRESS AND MYOCARDIAL PERFUSION

### HRV and Mental Stress Challenge {#FigureMimsViolin}

```{r}
mims_figures$violin
```

\newpage
\bland
\bfoot

### Distribution of HRV and Mental Stress Challenge {#TableMimsPaired}

```{r}
mims_tables$paired
```

\efoot
\eland
\newpage
\bfoot

### Distribution of HRV and MSIMI {#TableMimsHRV}

```{r}
mims_tables$hrv %>%
	tab_header(title = "HRV distribution by MSIMI", subtitle = "MIMS/MIPS cohorts") %>%
	tab_source_note("The distribution of HRV between those with MSIMI and those without. The HRV metric are stratified by phase of mental stress challenge. MSIMI = mental stress-induced myocardial ischemia, HRV = heart rate variability.")
```

\efoot
\newpage
\bfoot

### Depression and PTSD with Mental Stress Challenge {#TableMimsPsych}


```{r}
mims_reports$psych %>%
	tab_header(title = "Mental Stress Challenge HRV and Chronic Psychological Stress", subtitle = "MIMS/MIPS Cohorts") %>%
	tab_source_note("The association between HRV during mental stress challenge and the chronic psychological stressors of depression and PTSD are described. HRV = heart rate variability.")
```

\efoot
\newpage
\bland
\bfoot

### HRV and Chronic Mental Stress in Twins {#TableTwinsPsych}

```{r}
twins_reports$psych %>% 
	tab_header(title = "Morning HRV and Chronic Psychological Stress", subtitle = "Emory Twins Study") %>%
	tab_source_note("Depression is measured as a binary outcome with Beck Depression Inventory score > 14. PTSD = Post-Traumatic Stress Disorder, HRV = heart rate variability, LF = low frequency HRV, HF = high frequency HRV, VLF = very low frequency HRV, AC = acceleration capacity")
```

\efoot
\eland
\newpage
\bfoot

### Circadian HRV and Chronic Mental Stress {#TableTwinsCircPsych}

```{r}
twins_reports$psych_cosinor %>%
	tab_header(title = "Circadian HRV and Chronic Psychological Stress", subtitle = "Emory Twins Study") %>%
	tab_source_note("Depression is measured as a binary outcome with Beck Depression Inventory score > 14. The HRV metrics are measured over 24 hours using cosinor statistics.  PTSD = Post-Traumatic Stress Disorder, HRV = heart rate variability, LF = low frequency HRV, HF = high frequency HRV, VLF = very low frequency HRV, AC = acceleration capacity, MESOR = midline estimating statistic of rhythm, Amplitude = maximum distance from MESOR, Phi = shift of acrophase")
```

\efoot
\newpage
\bland
\bfoot

### Modeling Mental Stress-Induced Myocardial Ischemia and HRV {#TableMimsMSIMI}

```{r}
mims_reports$mpi %>%
	tab_header(title = "Mental Stress-Induced Myocardial Ischemia and HRV", subtitle = "MIMS/MIPS Cohorts") %>%
	tab_source_note("The association between the exposure of HRV with the finding of MSIMI is described. The HRV metric are stratified by phase of mental stress challenge. MSIMI = mental stress-induced myocardial ischemia, HRV = heart rate variability.")
```

\efoot
\eland
\newpage

## CLINICAL OUTCOMES

\bland
\bfoot

### Outcomes in MIMS/MIPS

```{r}
mims_reports$outcomes %>%
	tab_header(
		title = "Outcomes Analysis for Mental Stress and HRV",
		subtitle = "Traditional and Recurrent Event Models in MIMS/MIPS"
	) %>%
	tab_source_note("Estimates = HR (95% CI). Bolded terms signify p-value < 0.05. PWP = Prentice, Williams, and Peterson models, MSIMI = Mental Stress-Induced Myocardial Ischemia, LF = Low Frequency, HF = High Frequency, HRV = Heart Rate Variability")
```

\efoot
\eland
\newpage
\bland
\bfoot

### Outcomes in Twins {#TableTwinsOutcomes}

```{r}
twins_reports$outcomes %>%
	tab_header(title = "Clinical Outcomes by HRV", subtitle = "Emory Twins Study") %>%
	tab_source_note("Every unit increased in HRV had the associated hazard ratio (95% CI) for both overall and cardiovascular mortality. HRV = heart rate variability.")
```

\efoot
\eland


\begin{comment}
\end{comment}

<!--chapter:end:dissertation/08-appendix.Rmd-->

