---
title: |
  Stress Reactivity  
  Disturbances of the Neurocardiac Axis
subtitle: "Anish Sanjay Shah"
author: |
  A thesis submitted for the degree of  
  *Master of Science in Clinical Research*   
  Laney Graduate School  
  Emory University  
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: 
  bookdown::pdf_book:
    includes:
      in_header: preamble.tex
    lot: no
    lof: no
    toc: false
bibliography: ["../bibliography/Neurocardiology.bib", "../bibliography/Publications.bib"]
csl: "../bibliography/american-journal-of-epidemiology.csl"
link-citations: yes
documentclass: book
classoption: openany
colorlinks: yes
graphics: yes
geometry:
- top=1in
- bottom=1in
- left=1.5in
- right=1.5in
fontsize: 11pt
mainfont: Constantia
monofont: "Source Code Pro"
monofontoptions: "Scale=0.8"
header-includes:
  - \AtBeginDocument{\frontmatter}	
  - \usepackage{float}
  - \usepackage{longtable}
  - \usepackage{booktabs}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
  - \usepackage{graphicx}
---

```{r setup, global_options, include=FALSE}
# Knitr layout options
knitr::opts_chunk$set(
  cache = FALSE,
  eval = TRUE,
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  dpi = 300,
  dev = "png",
  options("scipen" = 0, "digits" = 2)
)
```

```{r flair_color}
library(flair)
darkblue <- "#0C2340"
gold <- "#B58500"
blue <- "#012169"
yellow <- "#F2A900"
teal <- "#487F84"
orange <- "#C35413"
red <- "#DA291C"
violet <- "#6558B1"
magenta <- "#C6007E"
lightblue <- "#008DBA"
black <- "#101820"
grey <- "#B1B3B3"
```

\begin{singlespace}
\tableofcontents
\end{singlespace}

# Acknowledgements

Foremost, I would like to thank my mentors and mentoring team. Dr. Shah has been a tremendous supporter of my academic growth, and has challenged me to continue to explore and learn, and create an area of academic excellence for myself. Dr. Alonso and Dr. Thames have all supported my efforts through direct mentorship and academic guidance. Dr. Quyyumi and Dr. Vaccarino have created a research environment that has laid the groundwork for these projects.

Second, I would like to thank the support staff that has allowed this research to unfold, from the EPICORE group with Nancy Murrah and Lucy Shallenberger, to the MSCR program with Cheryl Sroka and the outstanding MSCR instructors, including Dr. Matthew Magee.

Finally, I would like to thank my family for their support, including Michelle for her patience, and to Ripley and Harper for the always welcome distractions.

# Abbreviations

\bfoot
\vspace*{\fill}
{
\centering

```{r abbreviations}
tribble(
	~Term, ~Abbreviation,
	"HRV", "Heart Rate Variability",
	"CAD", "Coronary Artery Disease",
	"IHD", "Ischemic Heart Disease",
	"MACE", "Major Adverse Cardiovascular Events",
	"Biobank", "Emory Cardiovascular Biobank",
	"MIMS", "Myocardial Infarction and Mental Stress",
	"MIPS", "Mental Stress Ischemia Prognosis Study",
	"Twins", "Emory Twins Study",
	"PTSD", "Post-Traumatic Stress Disorder",
	"MI", "Myocardial Infarction",
	"MPI", "Myocardial Perfusion Imaging",
	"CFR", "Coronary Flow Reserve",
	"MSIMI", "Mental Stress-Induced Myocardial Ischemia",
	"PSIMI", "Physical Stress-Induced Myocardial Ischemia",
	"ANS", "Autonomic Nervous System"
) %>%
	arrange(Term) %>%
	gt() 
```

}
\vspace*{\fill}
\efoot
\clearpage

```{r, include=FALSE}
# Overview
tar_load(diagrams)

# Biobank
tar_load(biobank_tables)

# Twins
tar_load(twins_reports)
tar_load(twins_tables)
tar_load(twins_ecg)
tar_load(twins_clinical)
tar_load(twins_models)

# MIMS/MIPS
tar_load(mims_tables)
tar_load(mims_figures)
tar_load(mims_reports)
tar_load(mims_clinical)
tar_load(mims_models)
```
