# (PART\*) RESULTS {-}

# Clinical Characteristics

The study populations in the three cohorts are uniquely suited for these analyses. They are complementary in their description of cardiovascular disease, autonomic function, and psychological factors, and are described here.

The *Biobank* cohort, as described in Table \@ref(TableOneBiobank), had `r biobank_tables$one$N` participants, with a mean (95% CI) age in years of `r inline_text(biobank_tables$one, age)`. 
`r inline_text(biobank_tables$one, gend, level = Female)` were female, and `r inline_text(biobank_tables$one, race, level = "African American Black")` were Black. 
There were `r inline_text(biobank_tables$one, stenosis)` that had obstructive CAD on coronary angiography, and `r inline_text(biobank_tables$one, sad)` had depression.

```{r, include=FALSE}
mims <- 
	mims_clinical %>%
	select(c(age_bl, female_bl, race_bl, bmi, smk_now1, cath_cad70binary, starts_with("hx_"), rdr_msi_bl, rdr_psi2_bl, scid_depression_bl, scid_ptsd_bl)) %>%
	select(-c(hx_revasc_bl)) %>%
	tbl_summary(
		type = list(c(rdr_psi2_bl, scid_ptsd_bl, scid_depression_bl, female_bl) ~ "dichotomous", c(race_bl) ~ "categorical"),
		value = list(female_bl ~ "Female"),
		missing = "no"
	)
```

The *MIMS/MIPS* cohort had `r mims$N` participants.
The mean age was `r inline_text(mims, age_bl)`, `r inline_text(mims, female_bl)` were female, and `r inline_text(mims, race_bl, level = Black)` were Black.
`r inline_text(mims, cath_cad70binary)` had obstructive CAD.
`r inline_text(mims, scid_depression_bl)` had a diagnosis of depression, and `r inline_text(mims, scid_ptsd_bl)` had a diagnosis of PTSD. 
In this population, `r inline_text(mims, rdr_msi_bl, level = 2)` had MSIMI.
Additional breakdown by study group is described in \@ref(TableOneMims).

The *Twins* cohort, as described in both Table \@ref(TableOneTwins) and \@ref(TableTwinsHRV), had `r twins_tables$one$N` participants over 4 follow-up visits, with `r length(unique(twins_clinical$vetrid))` unique participants. 
The mean age was `r inline_text(twins_tables$one, age, column = THS1)` during the initial enrollment period, and was `r inline_text(twins_tables$one, age, column = ETSF)` during the final enrollment period. 
All participants were male, and `r prop.table(table(twins_clinical$race))[1]*100`% were White.
The average rate over the enrollment periods of abnormal MPI was `r prop.table(table(twins_clinical$pet_bin))[2]*100`%.
The average rate of PTSD was `r prop.table(table(twins_clinical$ptsd))[2]*100`% and the average rate of depression was `r prop.table(table(twins_clinical$sad_bin))[2]*100`%.

# Myocardial Ischemia

The relationship of autonomic dysfunction to CAD as measured by coronary angiography was assessed in the *Biobank* cohort. 
When comparing summary HRV metrics between those with obstructive CAD versus nonobstructive CAD, there were no significant differences between HRV distributions (\@ref(TableBiobankCAD)). 
When comparing those that had revascularization of the CAD and those that did not (\@ref(TableBiobankRevasc)), there was a difference seen in RR interval. 
Those that underwent revascularization had a mean (95% CI) RR interval of `r inline_text(biobank_tables$revasc, n_nmean, column = stat_2)`, while those that did not had a mean RR interval of `r inline_text(biobank_tables$revasc, n_nmean, column = stat_1)`.
There was a trend towards an increased *Dyx* in those that underwent revascularization (`r inline_text(biobank_tables$revasc, dyx, column = stat_2)`) than those that did not (`r inline_text(biobank_tables$revasc, dyx, column = stat_1)`).
No other HRV metrics were associated with revascularization.
To effect of the timing of revascularization on the subsequent changes in HRV acutely were assessed, as described in Table \@ref(TableBiobankTiming). 
No differences were seen between HRV before or after cardiac catheterization.

```{r, include=FALSE}
mims <-
	mims_models$equipment$ischemia %>%
	unnest(tidied) %>%
	filter(term != "(Intercept)") %>%
	filter(outcomes == "rdr_msi_bl")

hf <- mims[mims$term == "hf_stress",]
lf <- mims[mims$term == "lf_stress",]
```

The relationship of autonomic dysfunction to qualitative MPI was assessed using both mental stress and physical stress in the *MIMS/MIPS* cohorts.
ECG and HRV metrics did not have an association with abnormal MPI with combined mental and physical stress nor with physical stress.
Both lf HRV and LF HRV most prominently had an association with MSIMI, with stress HRV HRV having an odds ratio (OR) = `r hf$estimate` (95% CI `r hf$conf.low`, `r hf$conf.high`) and LF HRV having an OR = `r lf$estimate` (95% CI `r lf$conf.low`, `r lf$conf.high`). 
The other associations are described in Table \@ref(TableMimsStress).

```{r, include=FALSE}
twins <- twins_reports$mpi[["_data"]]
mpi <- twins[twins$outcomes == "Abnormal MPI" & twins$test_num == "Model 3",]
cfr <- twins[twins$outcomes == "Coronary Flow Reserve" & twins$test_num == "Model 3",]
```

This relationship between myocardial perfusion and autonomic dysfunction was further explored using quantitative MPI in the *Twins* cohort. 
Morning HRV at approximately 7 AM was predominately associated with coronary flow reserve, as described in Table \@ref(TableTwinsMPI). 
A change in 1 unit of LF HRV was associated with an `r cfr$lf_estimate` (95% CI `r cfr$lf_conf.low`, `r cfr$lf_conf.high`) in adjusted models. 
*Dyx* had an OR = `r mpi$dyx_estimate` (95% CI `r mpi$dyx_conf.low`, `r mpi$dyx_conf.high`) for abnormal MPI.

```{r, include=FALSE}
# Findings
twins <- twins_reports$ischemia_cosinor[["_data"]]
mpi <- twins[twins$outcomes == "Abnormal MPI", ]
cfr <- twins[twins$outcomes == "Coronary Flow Reserve",]
```

Within the *Twins*, diurnal HRV metrics were measured using cosinor analysis. 
The relationship of the MESOR, amplitude, and acrophase with abnormal MPI and coronary flow reserve were evaluated (Table \@ref(TableTwinsCircMPI)). 
The MESOR in particular showed a consistent relationship with coronary flow reserve, with a `r mpi$mesor_estimate[mpi$level == "Low Frequency HRV"]` (`r mpi$mesor_conf.low[mpi$level == "Low Frequency HRV"]`, `r mpi$mesor_conf.high[mpi$level == "Low Frequency HRV"]`) increase in every 1 unit increase in LF HRV, and a `r mpi$mesor_estimate[mpi$level == "Dyx"]` (`r mpi$mesor_conf.low[mpi$level == "Dyx"]`, `r mpi$mesor_conf.high[mpi$level == "Dyx"]`) increase for every 1 unit increase in *Dyx*.

# Psychological Stress and Myocardial Ischemia

Chronic psychological stressors were analyzed using all three cohorts. 
In the *Biobank* cohort, there were no significant differences seen in HRV by depressive symptoms as measured on the PHQ-9.

```{r, include=FALSE}
ptsd <- twins_reports$psych[["_data"]] %>%
	filter(test_num == "Model 3") %>%
	filter(outcomes == "PTSD")

sad <- twins_reports$psych[["_data"]] %>%
	filter(test_num == "Model 3") %>%
	filter(outcomes == "Depression")
```

In the *Twins*, early morning HRV was measured against both PTSD and depression. 
There was a significant relationship between HRV and both depression and PTSD as seen in Table \@ref(TableTwinsPsych).
In adjusted logistic models for PTSD, every 1 unit increase in HF HRV had an OR = `r ptsd$hf_estimate` (95% CI `r ptsd$hf_conf.low`, `r ptsd$hf_conf.high`), and LF HRV had an OR = `r ptsd$lf_estimate` (95% CI `r ptsd$lf_conf.low`, `r ptsd$lf_conf.high`).
In adjusted models logistic models for depression, every 1 unit of increase in VLF HRV had an OR = `r sad$vlf_estimate` (95% CI `r sad$vlf_conf.low`, `r sad$vlf_conf.high`).
*Dyx* and VLF HRV were not strongly associated with PTSD. 

```{r, include=FALSE}
ptsd <- twins_reports$psych_cosinor[["_data"]] %>%
	filter(outcomes == "PTSD") %>%
	filter(level == "Low Frequency HRV")

sad <- twins_reports$psych_cosinor[["_data"]] %>%
	filter(outcomes == "Depression") %>%
	filter(level == "Low Frequency HRV")
```

WHen using the diurnal HRV metrics, measured by cosinor analysis, significant relationships were seen with both depression and PTSD in the MESOR and amplitude (Table \@ref(TableTwinsCircPsych)).
For example, every 1 unit increase in the MESOR of LF HRV had an OR = `r ptsd$mesor_estimate` (95% `r ptsd$mesor_conf.low`, `r ptsd$mesor_conf.high`) and every 1 unit increase in the amplitude had an OR = `r ptsd$amp1_estimate` (95% `r ptsd$amp1_conf.low`, `r ptsd$amp1_conf.high`) for PTSD.
Every 1 unit increase in the MESOR of LF HRV had an OR = `r sad$mesor_estimate` (95% `r sad$mesor_conf.low`, `r sad$mesor_conf.high`) and every 1 unit increase in the amplitude had an OR = `r sad$amp1_estimate` (95% `r sad$amp1_conf.low`, `r sad$amp1_conf.high`) for depression.

Acute mental stress was also assessed primarily using the *MIMS/MIPS* cohorts. 
The distribution of HRV metrics based on the phase of acute mental stress challenge was evaluated, as seen in Figure \@ref(FigureMimsViolin). 
There were small differences between stress and rest HRV, as seen in Table \@ref(TableMimsPaired).
The difference in distribution of HRV was compared between those that had MSIMI and those that did not, as described in \@ref(TableMimsHRV).
There was a decrease in HRV in those with MSIMI compared to those without, except with heart rate. 

```{r, include=FALSE}
mims <- mims_reports$psych[["_data"]]
hr <- mims[mims$hrv == "hr" & mims$phase == "Rest", ]
lf <- mims[mims$hrv == "lf" & mims$phase == "Recovery", ]
```

The association between HRV during acute mental stress and chronic mental stress was also assessed (Table \@ref(TableMimsPsych)).
Every 10 beat/minute increase in resting heart rate had an OR = `r hr$scid_ptsd_bl_estimate` (95% CI `r hr$scid_ptsd_bl_conf.low`, `r hr$scid_ptsd_bl_conf.high`) for PTSD and an OR = `r hr$scid_depression_bl_estimate` (95% CI `r hr$scid_depression_bl_conf.low`, `r hr$scid_depression_bl_conf.high`) for depression. 
Every 1 unit increase in LF HRV during recovery had an OR = `r lf$scid_ptsd_bl_estimate` (95% CI `r lf$scid_ptsd_bl_conf.low`, `r lf$scid_ptsd_bl_conf.high`) for depression. 
No other HRV metrics were strongly associated.

```{r, include=FALSE}
mims <- mims_reports$mpi[["_data"]] %>%
	filter(test_num == "Model 5")
```

To assess the relationship of acute mental stress with myocardial perfusion abnormalities, the relationship between HRV and MSIMI was assessed. As seen in Table \@ref(TableMimsMSIMI), there was a robust association between LF and HF HRV during rest and stress with MSIMI. In fully adjusted models, including adjustment for both cardiovascular and psychological risk factors, every 1 unit increase in stress HF HRV had an OR = `r mims$hf_stress_estimate` (95% CI `r mims$hf_stress_conf.low`, `r mims$hf_stress_conf.high`) and stress LF HRV had an OR = `r mims$hf_stress_estimate` (95% CI `r mims$lf_stress_conf.low`, `r mims$lf_stress_conf.high`) for MSIMI.

# Clinical Outcomes

```{r, include=FALSE}
twins <- twins_reports$outcomes[["_data"]] %>%
	filter(outcome == "All Cause Mortality" & test_num == "Model 5")

circ <- twins_reports$outcomes_cosinor[["_data"]] %>%
	filter(outcome == "All Cause Mortality" & level == "Dyx")
```

Clinical outcome data was available in both the *Twins* and the *MIMS/MIPS* cohorts. 
With the *Twins*, early morning HRV showed a robust association with overall mortality and with cardiovascular disease, as seen in Table \@ref(TableTwinsOutcomes).
In fully adjusted models for overall mortality, *Dyx* and VLF HRV had the strongest association. 
With every 1 unit of increase in *Dyx*, there was a hazard ratio (HR) = `r twins$dyx_estimate` (95% CI `r twins$dyx_conf.low`, `r twins$dyx_conf.high`), and with every 1 unit increase in VLF HRV, there was a HR = `r twins$vlf_estimate` (95% CI `r twins$vlf_conf.low`, `r twins$vlf_conf.high`).
When evaluating the relationship of circadian changes in HRV and clinical outcomes, *Dyx* was a significant predictor of both overall and cardiovascular mortality.
The MESOR of *Dyx* had a HR = `r circ$mesor_estimate` (95% CI `r circ$mesor_conf.low`, `r circ$mesor_conf.high`) and the amplitude of *Dyx* had a HR = `r circ$amp1_estimate` (95% CI `r circ$amp1_conf.low`, `r circ$amp1_conf.high`).
Further relationships are outlined in \@ref(TableTwinsCircOutcomes).

```{r, include=FALSE}
mims <- mims_reports$outcomes[["_data"]] %>%
	filter(test_num == "Model 6")
lf <- mims %>% filter(hrv == "Stress Low Frequency HRV")
hf <- mims %>% filter(hrv == "Stress High Frequency HRV")
```

Using the *MIMS/MIPS* cohorts, stress HRV was compared with clinical outcomes. 
There was a robust relationship between stress HRV and overall mortality, cardiovascular mortality, and recurrent cardiovascular events as described in Table \@ref(TableMimsOutcomes).
In fully adjusted models for cardiovascular mortality, including adjustment for MSIMI, 1 unit increase in stress LF HRV had a HR = `r lf$cv_death_estimate` (95% CI `r lf$cv_death_conf.low`, `r lf$cv_death_conf.high`) and HF HRV had a HR = `r lf$cv_death_estimate` (95% CI `r lf$cv_death_conf.low`, `r lf$cv_death_conf.high`).
