---
title: Disturbances of the Neurocardiac Axis
subtitle: MSCR Thesis Overview
author:
  - Anish Shah, MD\textsuperscript{1}
institute:
  - \textsuperscript{1}Department of Epidemiology, Rollins School of Public Health,  Emory University, Atlanta, Georgia
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
 xaringan::moon_reader:
   css: [default, metropolis, metropolis-fonts]
   nature:
     titleSlideClass: ["center", "middle"]
---

```{r include = FALSE}
knitr::opts_chunk$set(
   fig.retina = 3,
   warning = FALSE,
   message = FALSE,
   echo = FALSE
)
```

# Introduction

### Thesis Overview

- Temporal relationship of ANS dysfunction with psychological and cardiac disease
   - **Acute**: the active stressor, such as mental stress or acute coronary reperfusion
   - **Diurnal**: both daily and circadian parameters and cross-sectional relationships
   - **Longitudinal**: effect of ANS function on clinical outcomes
   
- Neurocardiology assumes ANS is bidirectional bridge between heart and brain
   - Limitations in ability to non-invasively measure ANS function
   - Prevalent in many disease states (e.g. depression increases risk of CV disease)
   - ECG and signal-processing methods can help assess these changes in short-term and long-term recordings

---
# Background I

.center[
*Heart rate variability (HRV) can represent the current state of the ANS as a potential "biomarker" of neurocardiac health.*
]

.pull-left[
### Heart rate fluctuations

- Fluctuations in the HR are mediated by sympathetic (SNS) and parasympathetic (PNS) inputs to the sinoatrial node
- Rapid fluctuations in HR reflect vagal/PNS control
- Slower fluctuations in HR reflect SNS and PNS integartion, along with other influences
]

.pull-right[
### Relevant factors

- External influences such as activity, mental stress, etc
- Internal period rhythms also exist: RSA, baroreceptor reflexes, circadian rhythm, RAAS/neuroendocrine secretions, etc
]

---
# Background II

.pull-left[
## Psychological distress

- Acute stressors, such as a mental stress challenge, can lead to increased sympathetic activity and vagal withdrawal
- Psychological distress, such as psychiatric illness including depression and anxiety, can lead to changes in ANS function
- Psychosocial stress, such as chronic vital exhaustion and social isolation, can lead to chronic changes in ANS function
- However, ANS function may be protective by providing resilience to psychosocial stress
]

.pull-right[
## Cardiovascular disease

- Afferent pathways from cardiovascular disease can lead to both acute and chronic changes in ANS function
- Coronary occlusion, from perhaps myocardial infarction, leads to strong ANS response through afferent systems
- Chronic cardiovascular disease, can lead to a persistently elevated ANS response, e.g. resting tachycardia
- ANS dysfunction likely effects not only SA/AV nodal function, but myocardial inotropy, systemic vascular tone, and coronary resistance vessels 
]

---
# Background III

The background work supporting the feasability of the aims is as follows:

- ECG signal for Biobank, Twins, MIMS/MIPS has been generated previously

- HRV and perfusion data has been shown to be related in a circadian fashion [1]

- HRV and long-term outcomes has been shown to effect psychosocial factors and mortality/morbidity [2]

- Harmonic regression methods have been developed for analyzing diurnal data [3]

- Modeling package to assist with multiple outcomes [4]

.footnote[
[1] [Shah et al 2020](https://doi.org/10.1016/j.ijcard.2020.01.069)  
[2] Accepted manuscript in JAHA, presented at [AHA Epidemiology](https://www.ahajournals.org/doi/10.1161/circ.141.suppl_1.P542)  
[3] R Package called [card](https://cran.r-project.org/web/packages/card/index.html)  
[4] R package called [marksman](https://github.com/asshah4/marksman)
]

---
# Specific Aims Overview

Using available clinical data, will review the relationship of autonomic dysfunction with neurocardiac changes acutely, over the course of a day, and over the course of months to years.

**AIM #1: To establish the relationship of coronary artery perfusion and reperfusion (through revascularization) with autonomic nervous system function. **
	
**AIM #2: To establish the effect of autonomic function on the relationship between psychological distress, both acute and chronic, and abnormal myocardial perfusion.** 

**AIM #3: To examine the effect of stress reactivity on risk prediction of long-term events.**

**AIM #4: To explore the utility of advanced harmonic regression modeling and feature reduction methods in ECG signal processing techniques in statistical analysis.** 

---
# Aim I

**AIM #1: To establish the relationship of coronary artery perfusion and reperfusion (through revascularization) with autonomic nervous system function. ** *In patients with obstructive CAD, reperfusion will lead to increased HRV afterwards as compared to patients without obstructive CAD undergoing cardiac catheterization. The baseline HRV will have a inverse relationship with psychological stress (measured by BDI).*

.pull-left[
- Stable CAD + psychological stress has $\uparrow$ CV mortality, potential through ANS
- Unknown how reperfusion affects ANS in stable CAD
- ECG data (HRV) can be collected before/after reperfusion
- Control for intra-procedural variables using *nonobstructive* CAD patients
- **Biobank** cohort
]

.pull-right[
```{r}
dagify(
   hrv ~ cad + pci + psych,
   pci ~ cad,
   labels = c(
      "hrv" = "HRV",
      "pci" = "Revascularization",
      "psych" = "Psychological Stress",
      "cad" = "CAD"
   )) %>%
   ggdag(text = FALSE, use_labels = "label") +
   theme_dag()
```
]

---
# Aim II

**AIM #2: To establish the effect of autonomic function on the relationship between psychological distress, both acute and chronic, and abnormal myocardial perfusion.** *Acute mental stress will lead to abnormalities in HRV. HRV will be associated with chronic psychological distress, including depression/anxiety and allostatic load. The abnormalities will be most prominently seen in changes to circadian rhythm of HRV. Abnormalities in HRV will associate with increased risk of abnormal MPI and of MSIMI.*

.pull-left[
- Acute mental stress can lead to inducible myocardial ischemia (MSIMI)
- Chronic psychological stress effects ANS function (circadian rhythm, coronary flow reserve) 
- Acute and chronic psychological stress show $\uparrow$ risk of cardiovascular mortality/morbidity
- **Twins** and **MIMS/MIPS** cohorts
]

.pull-right[
```{r}
dagify(
   mpi ~ hrv,
   hrv ~ ms + ps + psych,
   ms ~ psych,
   mpi ~ ms + ps,
   circ ~ psych + hrv,
   labels = c(
      "hrv" = "HRV",
      "psych" = "Psychological Stress",
      "mpi" = "Myocardial Ischemia",
      "ms" = "Mental Stress",
      "ps" = "Physical Stress",
      "circ" = "Circadian Changes"
   )) %>%
   ggdag(text = FALSE, use_labels = "label", layout = "tree") +
   theme_dag()
```
]

---
# Aim III

**AIM #3: To examine the effect of stress reactivity on risk prediction of long-term events.** *Abnormal HRV in response to mental stress will lead to increased cardiovascular disease events. Abnormal circadian changes in HRV will associate with increased cardiovascular disease events and susceptibility to incident psychological distress.*

.pull-left[
- Stress reactivity defined as abnormal ANS response to *everyday* stressor
- Includes HRV responses to... 
   - mental stress 
   - physical stress
   - abnormal coronary flow reserve
   - circadian disruption
- Abnormal stress reactivity has few known associations with outcomes
- **Twins** and **MIMS/MIPS** cohorts
]

.pull-right[
```{r}
dagify(
   outcomes ~ ms + cfr + circ + hrv,
   ms ~ hrv,
   cfr ~ hrv,
   circ ~ hrv,
   labels = c(
      "hrv" = "HRV",
      "ms" = "Mental Stress",
      "cfr" = "Abnormal Coronary Flow Reserve",
      "circ" = "Circadian Changes",
      "outcomes" = "Outcomes"
   )
) %>%
   ggdag(text = FALSE, use_labels = "label", layout = "nicely") +
   theme_dag()
```
]

---
# Aim IV

**AIM #4: To explore the utility of advanced harmonic regression modeling and feature reduction methods in ECG signal processing techniques in statistical analysis.** *Circadian changes, measured through cosinor metrics, will have greater predictivity of disturbances of the neurocardiac axis (both psychological distress and myocardial perfusion abnormalities) than that of 24-hour HRV summary statistics.*

.pull-left[
- Harmonic regression is powerful tool to assess autoregressive data
- Multiple periodicities exist in human diurnal variation of ANS function
- Statistical models that incorporate both population and multiple components will have greater granularity for analyzing circadian rhythms
]

.pull-right[
```{r}
data(twins)
model <- cosinor(rDYX ~ hour, twins, tau = c(24, 12))
ggcosinor(model)
```
]

---
# Study Design I

### Study Overview and Data Collection 

*Emory Biobank*: Psychological factors and HRV collected on patients undergoing coronary angiography. Currently enrolling rare patients. [1]

*Emory Twins Study*: Veteran twins present for clinical profiling, Holter monitoring, and quantitative myocardial perfusion imaging. Follow-up for outcomes. Enrollment and follow-up complete.

*MIMS/MIPS*: Mental stress study looking at inducible myocardial ischemia and clinical outcomes. HRV collected before, during, and after mental stressors. Enrollment and follow-up complete.

*ARIC*: Large epidemiology dataset with >20+ years of patient follow-up, including short-term HRV and several time points. Enrollment and follow-up complete.

.footnote[
[1] _Due to COVID-19, the datasets are reduced in size, and alternative data sources were used_.
]

---
# Study Design II

### ECG Measures

Within the several datasets that will be used, ECG is collected in a raw format in the Emory Biobank, Twins, MIMS/MIPS and has already been processed to different HRV signals, including morphology changes. In ARIC, this data has been pre-processed to HRV.

### Coronary/cardiovascular measures

Within the Biobank, coronary angiography data is collected, including intervention. Within Twins, quantitative PET data has been collected and analyzed (showing myocardial perfusion data and coronary flow reserve).

### Clinical/psychological measures

For both Biobank (PHQ-9), MIMS/MIPS (BDI), and Twins (BDI), ARIC (vital exhaustion), extensive clinical and psychological profiling is available. MIMS/MIPS in addition includes a mental-stress challenge during their study visit.

---
# Analytical Plans I

## Acute Changes

1. Assess pre/post reperfusion HRV, with controls being patients with no intervention. Further adjustment for coronary stenosis. (**Biobank**)
   - Coronary angiography scores and pre/post flow rates after intervention 
   - Include effect of depression (PHQ-9) on HRV (additional covariate and/or strata)

1. Effect of acute mental stress on HRV (**MIMS/MIPS**)
   - Effect of acute mental stress and immediate changes in HRV
   - Assess relationship of mental stress and HRV "recovery" to baseline
   - Relationship between those with and without MSIMI (mental stress-induced myocardial ischemia) and HRV
   
---
# Analytical Plans II

## Diurnal Changes

1. Create and implement a harmonic regression model with multiple components to assess ECG features 
   - Assess implementation in 24 hour datasets (**Biobank** and **Twins**)
   - Assess implementation in >24 hour datasets (**Twins** and **MIMS/MIPS**)
   - Assess implementation in <12 hour datasets (**MIMS/MIPS**)
   
1. Using cosinor regression to assess relationship of ECG with...
   - Myocardial perfusion abnormalities (**Twins**)
   - Depression via BDI and PHQ-9 (**Twins** and **Biobank**)
   - MSIMI (**MIMS/MIPS**)

	
---
# Analytical Plans III

## Longitudinal changes

1. Model relationship of circadian/cosinor ECG with future events
   - With adjustment for clinical/psychological covariates (**Twins**)
   - With adjustment/stratification by MSIMI (**MIMS/MIPS**)
   
1. Examine relationship of changes in HRV with long-term outcomes
   - Psychosocial stress including vital exhaustion, social isolation, and anger proneness (**ARIC**)

---
# Results and Conclusions

The analyses are in various stages of work, outlined here:

- Twins data has published analysis on circadian patterns on myocardial perfusion, longitudinal outcomes and relationship to depression are pending
- Biobank data has been collected, but comparative analyses has not been completed
- MIMS/MIPS data has been collected, and preliminary analyses have begun
- ARIC data and longitudinal outcomes have been analyzed, with association of mortality and HRV as an additive effect to psychosocial stress
