---
title: Disturbances of the Neurocardiac Axis
subtitle: MSCR Thesis Overview
author:
  - Anish Shah, MD\textsuperscript{1}
institute:
  - \textsuperscript{1}Department of Epidemiology, Rollins School of Public Health,  Emory University, Atlanta, Georgia
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
 xaringan::moon_reader:
   css: [default, metropolis, metropolis-fonts]
   self_contained: true
   nature:
     titleSlideClass: ["center", "middle"]
---

```{r include = FALSE}
knitr::opts_chunk$set(
   fig.retina = 3,
   warning = FALSE,
   message = FALSE,
   echo = FALSE
)

# While working on it
#library(ggdag)
#library(tidyverse)
#library(card)
```

# Introduction

Autonomic dysfunction is associated with both psychological abnormalities and cardiac disease. 

Manifestations:

- Inappropriate physiological response to acute stress (e.g. excessive heart rate increase)
- Changes in diurnal/circadian patterns of physiological systems (e.g. sleep-wake disturbances, sudden death in early mornings)
- Interaction with existing cardiovascular and psychiatric disease (e.g. increased mortality in comorbid depression and coronary artery disease)

Measurements:

- Measuring this bi-directional bridge between heart and brain is limited
- Non-invasive measures including changes in sinoatrial node function (a marker of sympathovagal balance), described as heart rate variability.
- Electrocardiogram data and signal-processing techniques can make assessments of these autonomic changes

---
# Abbreviations

| Abbreviation | Description |
| - | - |
| HRV | Heart Rate Variability |
| CAD | Coronary Artery Disease | 
| MACE | Major Adverse Cardiovascular Events |
| MPI | Myocardial Perfusion Imaging |
| CFR | Coronary Flow Reserve |
| MSIMI | Mental Stress-Induced Myocardial Ischemia |
| ANS | Autonomic Nervous System |
| $\Psi$ | Psychological/Psychiatric |
| BDI | Beck's Depression Inventory |
| SCID | Structured Clinical Interview for DSM-IV | 
| PHQ-9 | Patient Health Questionnaire, 9-items |

---
# Background I

.center[
*Heart rate variability (HRV) can represent the current state of the ANS as a potential "biomarker" of neurocardiac health.*
]

.pull-left[
### Heart rate fluctuations

- Fluctuations in the HR are mediated by sympathetic (SNS) and parasympathetic (PNS) inputs to the sinoatrial node
- Rapid fluctuations in HR reflect vagal/PNS control
- Slower fluctuations in HR reflect SNS and PNS integartion, along with other influences
]

.pull-right[
### Relevant factors

- External influences such as activity, mental stress, etc
- Internal period rhythms also exist: RSA, baroreceptor reflexes, circadian rhythm, RAAS/neuroendocrine secretions, etc
]

---
# Background II

.pull-left[
## Psychological distress

- Acute stressors, such as a mental stress challenge, can lead to increased sympathetic activity and vagal withdrawal
- Psychological distress, such as psychiatric illness including depression and anxiety, can lead to changes in ANS function
- Psychosocial stress, such as chronic vital exhaustion and social isolation, can lead to chronic changes in ANS function
- However, ANS function may be protective by providing resilience to psychosocial stress
]

.pull-right[
## Cardiovascular disease

- Afferent pathways from cardiovascular disease can lead to both acute and chronic changes in ANS function
- Coronary occlusion, from perhaps myocardial infarction, leads to strong ANS response through afferent systems
- Chronic cardiovascular disease, can lead to a persistently elevated ANS response, e.g. resting tachycardia
- ANS dysfunction likely effects not only SA/AV nodal function, but myocardial inotropy, systemic vascular tone, and coronary resistance vessels 
]


---
# Specific Aims Overview

1. **To examine the cross-sectional relationship between ANS function and coronary artery reperfusion (through revascularization).** 

1. **To determine the if ANS dysfunction mediates the relationship of acute/chronic $\Psi$ distress and CAD, measured by MPI and cardiac catheterization.**  

1. **To explore the association of abnormal stress reactivity on increased long-term MACE.**   

1. **To explore the utility of advanced harmonic regression modeling and feature reduction methods in ECG signal processing techniques in statistical analysis.**  

---
# Study Design I

### Emory Biobank

Cross-sectional design. Patients undergoing cardiac catheterization at EUH are enrolled, excluding those with acute coronary syndromes. Psychological and clinical profiling is done prior to catheterization. ECG data is collected before, during, and after catheterization. [1]

- ECG: Using single-lead patch data (VivaLNK), RR intervals collected throughout procedure
- CAD: Left heart catheterization with coronary angiography performed. If coronary disease was significant, PCI (revascularization) was performed.
- $\Psi$: Depressive symptom burden collected via PHQ-9 (Patient Health Questionnaire) 

.footnote[
[1] Due to COVID-19, the datasets are reduced in size, and alternative data sources were used.
]

---
# Study Design II

### Emory Twins Study

Observational/cross-sectional design. Veteran twins were enrolled for quantitative MPI. Patients were flown to study site at EUH and monitored for ~30 hours, with controlled schedules. Clinical characteristics and extensive psychological profiling performed. Follow-up data for outcomes collected. 

- ECG: 3-lead to 12-lead Holter monitors were placed. Data transformed to HRV from raw GE Seer format
- CAD: Quantitative MPI performed on all patients, with coronary flow reserve measured
- $\Psi$: Depression data collected using BDI, and anxiety/post-traumatic stress disorder collected by SCID
- Follow-up: Adjudicated outcome events (MACE, hospitalization, etc) were collected

---
# Study Design III

### MIMS/MIPS

Observational/cross-sectional design. Patients after recent catheterization for chest pain were enrolled, primarily Grady and EUH. They returned for first visit within 6 months of event. Mental stress and physical stress performed, followed by quantitative MPI. ECG collected before, during, and after mental stress challenge. Follow-up data for clinical outcomes collected.

- ECG: 1-lead to 12-lead ECG monitors were placed. Data transformed to HRV using MATLAB
- CAD: Stress test performed initially using either mental/physical stress. Quantitative MPI performed on all patients, with coronary flow reserve measured
- $\Psi$: Depression and anxiety data collected using SCID
- Follow-up: Adjudicated outcome events (MACE, hospitalization, etc) were collected

---
# Aim I

**AIM #1: To examine the cross-sectional relationship between ANS function and coronary artery reperfusion (through revascularization).** 
*Hypothesis: In obstructive CAD, reperfusion of the coronary arteries will associate with increased HRV compared to those without reperfusion.*

.pull-left[
- **Biobank** cohort receive continuous ECG monitoring and cardiac catheterization
- Some receive revascularization of coronary arteries
- Unknown how reperfusion affects ANS in stable CAD (potential confounder)
- Can control for peri-operative unknowns with *nonobstructive* CAD patients undergoing cath
]

.pull-right[
```{r}
dg <- dagify(
   hrv ~ cad,
   pci ~ cad,
   hrv ~ pci,
   labels = c(
      "hrv" = "HRV",
      "pci" = "Revascularization",
      "cad" = "CAD"
   ),
   exposure = "pci",
   outcome = "hrv"
)

tidy_dagitty(dg, layout = "nicely") %>%
	node_status() %>%
	ggplot(aes(x = x, y = y, xend = xend, yend = yend, color = status)) +
	geom_dag_point() +
	geom_dag_edges(arrow_directed = grid::arrow(length = grid::unit(10, "pt"), type = "closed")) +
	geom_dag_label_repel(aes(label = label)) + 
   scale_color_hue(breaks = c("exposure", "outcome"), l = 40, c = 35) + 
   guides(color = guide_legend(override.aes = list(fill = NA ))) + 
   theme(
      legend.position = "bottom",
      legend.box.background = element_rect(fill = "#FAFAFA", colour = NA),
      legend.title = element_blank(),
      legend.key = element_rect(fill = "#FAFAFA", colour = NA),
      legend.background = element_rect(fill = "#FAFAFA", colour = NA),
      panel.background = element_rect(fill = "#FAFAFA", colour = NA),
      plot.background = element_rect(fill = "#FAFAFA", colour = NA),
      axis.title = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank()
   )

```
]

---
# Aim I: Analytical Plan

### Methods

- Identify timing of PCI/revascularization during ECG recording
- Measure amount of coronary stenosis using Gensini score for CAD 

### Statistics

- HRV pre/post PCI using *Paired T-Test*
- HRV post-PCI to controls (no revascularization) using *ANOVA*

---
# Aim II

**AIM #2: To determine the if ANS dysfunction mediates the relationship of acute/chronic $\Psi$ distress and CAD, measured by MPI and cardiac catheterization.**  
*Hypothesis: 1) Acute/chronic stress will be associated with abnormal HRV. 2) HRV will be associated with abnormal MPI and obstructive CAD. 3) The association between stress and abnormal myocardial perfusion will be mediated by HRV.*

.pull-left[
- **Twins** has data on chronic stress and MPI 
- **MIMS/MIPS** has acute stress (mental stress challenge) and MPI
- **Biobank** has chronic stress and cardiac catheterization
- All cohorts: continuous ECG/HRV
- Acute/chronic stress are associated with myocardial ischemia
- Acute and chronic stress are associated with ANS dysfunction
]

.pull-right[
```{r}
dg <- dagify(
   mpi ~ hrv + ms + psych,
   hrv ~ psych + ms,
   labels = c(
      "hrv" = "HRV",
      "psych" = "Chronic Stress",
      "mpi" = "Myocardial Ischemia",
      "ms" = "Acute Stress"
   ),
   exposure = c("ms", "psych"),
   outcome = "mpi"
)

tidy_dagitty(dg, layout = "nicely") %>%
	node_status() %>%
	ggplot(aes(x = x, y = y, xend = xend, yend = yend, color = status)) +
	geom_dag_point() +
	geom_dag_edges(arrow_directed = grid::arrow(length = grid::unit(10, "pt"), type = "closed")) +
	geom_dag_label_repel(aes(label = label)) + 
   scale_color_hue(breaks = c("exposure", "outcome"), l = 40, c = 35) + 
   guides(color = guide_legend(override.aes = list(fill = NA ))) + 
   theme(
      legend.position = "bottom",
      legend.box.background = element_rect(fill = "#FAFAFA", colour = NA),
      legend.title = element_blank(),
      legend.key = element_rect(fill = "#FAFAFA", colour = NA),
      legend.background = element_rect(fill = "#FAFAFA", colour = NA),
      panel.background = element_rect(fill = "#FAFAFA", colour = NA),
      plot.background = element_rect(fill = "#FAFAFA", colour = NA),
      axis.title = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank()
   )
```
]

---
# Aim II: Analytical Plan

### Methods

- Measure HRV after acute stress events
- Measure abnormalities in MPI, including CFR
- Quantify markers of chronic $\Psi$ stress (depression, anxiety)

### Statistics

- HRV before and after acute stress using *Paired T-test* (MIMS/MIPS)
- HRV after acute stress compared to control using *ANOVA* (MIMS/MIPS)
- CAD/MPI and HRV using *logistic regression*, with adjustment for $\Psi$ covariates (all)
- Mediation of acute/chronic stress with CAD/MPI by HRV using *mediation analysis* (all)

---
# Aim III

**AIM #3: To explore the association of abnormal stress reactivity on increased MACE.**   
*Hypothesis: 1) Abnormal stress reactivity, defined by abnormal HRV response to acute/chronic stress, will associate with an increased HR for MACE as compared to normal stress reactivity responses. 2) Abnormal stress reactivity will show additive effect to the increased HR from abnormal MPI.*

.pull-left[
- In **Twins** and **MIMS/MIPS** cohort: 
   - continuous ECG/HRV
   - acute/chronic mental stress
   - quantified MPI / CFR data
   - longitudinal outcomes including MACE
- Stress reactivity defined as:
   - $\downarrow$ HRV + acute $\Psi$ distress
   - $\downarrow$ HRV + chronic $\Psi$ distress
   - Circadian disruption
]

.pull-right[
```{r}
dg <- dagify(
   outcomes ~ stress,
   outcomes ~ mpi,
   mpi ~ stress,
   stress ~ hrv + psych + ms + circ,
   labels = c(
      "stress" = "Stress Reactivity",
      "mpi" = "Myocardial Ischemia",
      "outcomes" = "Outcomes",
      "hrv" = "HRV",
      "ms" = "Acute Stress",
      "psych" = "Chronic Stress", 
      "circ" = "Circadian Changes"
   ),
   exposure = c("stress"),
   outcome = "outcomes"
) 

tidy_dagitty(dg, layout = "nicely") %>%
	node_status() %>%
	ggplot(aes(x = x, y = y, xend = xend, yend = yend, color = status)) +
	geom_dag_point() +
	geom_dag_edges(arrow_directed = grid::arrow(length = grid::unit(10, "pt"), type = "closed")) +
	geom_dag_label_repel(aes(label = label)) + 
   scale_color_hue(breaks = c("exposure", "outcome"), l = 40, c = 35) + 
   guides(color = guide_legend(override.aes = list(fill = NA))) +
   theme(
      legend.position = "bottom",
      legend.box.background = element_rect(fill = "#FAFAFA", colour = NA),
      legend.title = element_blank(),
      legend.key = element_rect(fill = "#FAFAFA", colour = NA),
      legend.background = element_rect(fill = "#FAFAFA", colour = NA),
      panel.background = element_rect(fill = "#FAFAFA", colour = NA),
      plot.background = element_rect(fill = "#FAFAFA", colour = NA),
      axis.title = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank()
   )
```
]

---
# Aim III: Analytical Plan

### Methods

- Create follow-up data in recurrent event format from both **Twins** and **MIMS/MIPS**
- Create definitions for stress reactivity by categorizing those with abnormal HRV and acute and chronic stress

### Statistics

- *Cox Proportional Hazard Models* (all)
   - MACE and stress reactivity
   - MACE and abnormal MPI 
   - Adjust by including both MPI and stress reactivity in models

---
# Aim IV

**AIM #4: To explore the utility of advanced harmonic regression modeling and feature reduction methods in ECG signal processing techniques in statistical analysis.**  
*Hypothesis: Circadian changes in continuous HRV, measured through cosinor metrics, will have stronger associations with $\Psi$ distress and abnormal myocardial perfusion then summary HRV measurements.* 

.pull-left[
- Uses ECG data from **Biobank**, **Twins**, and **MIMS/MIPS** to obtain cosinor metrics
- Methods-focused aim to develop appropriate modeling/statistical technique for analysis
]

.pull-right[
```{r}
data(twins)
model <- cosinor(rDYX ~ hour, twins, tau = c(24, 12))
ggcosinor(model) +
   theme(
      legend.box.background = element_rect(fill = "#FAFAFA", colour = NA),
      legend.title = element_blank(),
      legend.key = element_rect(fill = "#FAFAFA", colour = NA),
      legend.background = element_rect(fill = "#FAFAFA", colour = NA),
      panel.background = element_rect(fill = "#FAFAFA", colour = NA),
      plot.background = element_rect(fill = "#FAFAFA", colour = NA),
      axis.title = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank()
   )
```
]

---
# Aim IV: Analytical Plan

### Methods

- Create harmonic regression analysis tools in `R`
- Create multiple-component harmonic regression tools in `R`

### Statistics

- *Cosinor Regression*
   - HRV
   - Adjust/compare by psychiatric disease 
   - Adjust/compare by abnormal MPI
- Compare fit between *cosinor* and *linear* regression models

---
# Status, Results and Conclusions

The analyses are in various stages of work, outlined here:

- Twins data has published analysis on circadian patterns on myocardial perfusion [1], longitudinal outcomes and relationship to depression are pending
- ARIC data and longitudinal outcomes have been analyzed, with association of mortality and HRV as an additive effect to psychosocial stress [2]
- Cosinor and circadian analysis tools have been created [3]
- Thesis modeling approach has been designed [4]
- Biobank data has been collected, but comparative analyses has not been completed
- MIMS/MIPS data has been collected, and preliminary analyses have begun

.footnote[
[1] [Shah et al 2020](https://doi.org/10.1016/j.ijcard.2020.01.069)  
[2] Accepted manuscript in JAHA, presented at [AHA Epidemiology](https://www.ahajournals.org/doi/10.1161/circ.141.suppl_1.P542)  
[3] R Package called [card](https://cran.r-project.org/web/packages/card/index.html)  
[4] R package called [octomod](https://github.com/asshah4/octomod)
]
